Query Optimization for Dynamic Imputation

### 2017-05-15 paper/rebuttal.tex

\begin{document}
\section{Introduction}
The authors would like to thank the reviewers for their helpful and insightful comments.
We have taken steps to address each comment and we believe that the paper is significantly improved
as a result. In this document, we give a point-by-point response to the comments from each reviewer.

In the rebuttal, the reviewer's original comments are in black and our responses are in \resp{blue}.
In our draft, new or significantly changed text is highlighted in {\color{blue}blue}.

\section{Overall changes}
\begin{itemize}[leftmargin=*]
\item Redundant figures, excessive whitespace, formatting.

  \resp{Several reviewers commented on the amount of whitespace, so we have compressed many of the figures and removed whitespace to make room for additional material.}

\item \resp{
    We have improved our system since the paper was submitted by adding imputation for categorical variables.
    This has a minor impact on the experimental results.
    It improves the ability of the regression tree model to perform imputation by making these variables available to the model. Note that this has a small impact on the results presented in the experiments section, relative
    to the original version submitted.
  }

\item \resp{We extended the original paper with two additional popular imputation techniques: mean value imputation and hot-deck imputation.

We added a section discussing how to construct the imputation-related functions used by our cost model, which represent the API for a new imputation operator,
for these two new techniques and the reference technique used in the original submission (\Cref{sec:imputation}). }

@@

\item \resp{We extended experiments to include the new operators and provide more details on results for $\alpha$ (\Cref{sec:results}). We also added the Pareto frontiers for various plans (\Cref{fig:pareto-frontiers} ) and
explored planning times for more complex queries.}

\item \resp{We included a new section that provides a roadmap for implementing \ProjectName{} on top of an existing database (\Cref{sec:implementation}).}
\end{itemize}

@@

  \section{Reviewer 1}
  \begin{itemize}[leftmargin=*]
  \item W2. The paper completely obscures the ML technique being used: (1) It would be nice to have some discussion of the complete API for implementing an ML technique, as well as how one might implement a cost-estimator for one or two examples. (2) It would be helpful to see the impact of a few different approaches (e.g., Tree-based, Interpolation, Regression, FD-based) on performance and accuracy. (3) Given that the ML time estimator is being treated as a black box, it would be useful to see how sensitive the plan selector is to inaccurate estimates.

  \resp{
    We extended our evaluation to include two commonly used imputation methods: mean value imputation and hot-deck imputation.
    These methods are in addition to the original imputation method (CART), which is based on regression trees.
    The regression tree model incorporates correlations between attributes, so should be able to take advantage of any functional dependencies to make better predictions.
    Interpolation-based imputation is useful for data where the ordering of tuples is significant, like time series, but is not well suited for survey data where the ordering of tuples is arbitrary.

@@

    Our optimizer (like any query optimizer) is sensitive to the quality of its estimates.
    We believe that the cost functions we have chosen lead to good results overall, but there may be cases where our estimates lead to bad outcomes.
    It is an interesting and difficult problem to accurately estimate the quality of a regression model before running the model.
    Any work in this area could be used immediately to improve our algorithm.
  }

  \item W3. Reducing runtime to 10s is excellent, but comes at a very serious accuracy cost (3-5x worse SMAPE). Worse still, there's no way for the user to differentiate between the cases of Q4 and Q8 at runtime. 24\% error relative to direct imputation is huge.

    \resp{
      The outsized error in Q8 is due to a drop operator placed very early on in the plan-- this is a result of optimizing for performance
      and the fact that this data set has many missing values for relevant attributes (77.50\% for \texttt{studentdebtowe}). This produced significantly smaller sample sizes for various
      groups, which skewed the average calculation for each group significantly. For example, the average student debt for subjects with no high school degree was almost 83\% off from the estimate
      using the base table imputation, but if we look at the underlying population the former was calculated with 6 people, while the latter was with 258 people.

      Note that after improving the categorical variable imputation in \ProjectName{}, the error for $\alpha=1.0$
      with the reference imputation method is now slightly lower in Q8 (19.99\%).

@@

      We can trivially extend \ProjectName{} to warn users when the query chosen for execution has a high $\textsc{Penalty}$ estimate, along with the number of tuples that have
      been dropped (after execution), so that situations such as those observed in Q8 can be identified by the user. We've added this detail to the paper (\Cref{sec:results}).
    }

  \item Although it doesn't consider cost or the drop operator, a similar technique appears in Yang et. al. "Lenses: An On-Demand Approach to ETL." A comparison would be nice.

   \resp{
     Yang's work is definitely related, and we've added it to our related work section.
   Their system helps an analyst prioritize cleaning tasks for a particular query.
   It uses pre-trained regression models as one of a number of repair strategies for missing data rather than creating a query plan which incorporates the training and running of such models (\Cref{subsec:missing-value}).
   }

@@

  \item S2: Why no unions?

    \resp{We skipped unions for brevity. There is nothing fundamental about our algorithm which makes unions difficult to implement.}

  \item S3.4.1, Definition 2: On my first read-through, I was puzzled about why there was no plan with zero-loss. It would help to clarify that Loss is not about tuple loss, but rather a heuristic penalty function.

    \resp{We have clarified this in the paper (\Cref{sec:cost-model}) by emphasizing that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense and renaming it to $\textsc{Penalty}$.}

@@

  \item S3.4.3: An extremely nice emergent property from the L-value function for the imputation operator seems to be that the operator will be moved to a position in the plan where it has access to the maximum training data. I am slightly concerned that by focusing entirely on size, this might lead to skew in the training set. For example, consider a one-to-many join \verb/µ(R) |X| S/. Your optimization algorithm would likely wish to move imputation up: \verb/µ(R |X| S)/. Yes, there is more training data, but if the fanout for each tuple is skewed, some tuples in R will become over-represented in the join result. A similar observation holds for selection. For example take your Q8, where the conditions "studentdebtowe $>$ 0 AND schooldegree $\geq$ 0" presumably indicate invalid data that should not be used for training. (is this perhaps why you have so much error for full imputation on Q8?)

    \resp{This is a reasonable concern. Though often an appropriate proxy, optimizing for the maximum training set is flawed when the query constructs a large, skewed data set. The problem seems to be particular to queries which generate large intermediate tables. In some cases, as when joining two datasets, the intermediate table is a good target for training because it exposes correlations which would be hidden in the separate tables. In other cases, this intermediate table is large and skewed, which causes problems for imputation.

@@

  \item S3.4.4: It appears that Time() assumes that plans are IO-bound. This is a perfectly reasonable assumption, but should be stated explicitly, as datasets that fit in memory will require a different cost model.

    \resp{Agreed. Our database stores tables in files, so our plans are I/O bound, but a database which uses an in-memory representation will need a different Time() function. We have clarified this in the paper (\Cref{sec:query-runtime}).}

  \item S3.4.4: "Drop is a special case of a sequential scan, and its time complexity can be expressed as a function of the number of heap pages read and the I/O cost per page." seems to suggest that drop is implemented as a blocking operator. Why not just take the same approach as selection and treat it as having negligible cost?

    \resp{We mistakenly suggested in the original submission that we treated \textit{Drop} as a blocking operator.
    In our implementation, it has $0$ cost, and we have updated the text to reflect that (\Cref{sec:query-runtime}).}

@@

  \item S4.4.1: Imputation over the raw data is being used as ground truth. Is there any reason to suspect that this is the correct thing to do? Perhaps joining multiple tables together exposes new attributes that makes imputation *more* reliable? At the very least, it would be useful to understand why there is so much error in $\alpha=0$ on Q2 and Q8.

    \resp{For our purposes, imputation over the raw data is a reasonable ground truth. This
      is because the analyst, in practice, should be considering the tradeoffs on quality
      and performance of using ImputeDB versus a production imputation system on the raw
      data. Unfortunately, they would never have access to real ground truth in such a
      scenario, and moreover, finding ground truth for use even in an empirical evaluation
      of imputation methods can be challenging because artificially dirtying the data
      often does not capture the complex reasons for missingness in real world datasets.
      We updated the discussion in \Cref{sec:results} to emphasize this point.}

  \item There are a number of opportunities for trimming in the figures, tables, and algorithms: Algorithms 1-3 could be compacted a bunch, Table 1 is full of whitespace (e.g., make it a 3-column {figure*}), Tables 1-2 could also be summarized as CDFs, Figure 6 is full of whitespace, Table 3 could be put queries on a single row (e.g., Q1 would easily fit on 1 line), and the references have a lot of redundant information (URLs, for example).

    \resp{We have compacted the algorithms and tables. Note that \Cref{table:nhanes-description, table:fcc-description} 1-2 show the percentage of missing values for \emph{each attribute}, not as an overall percentage, so cannot be summarized as CDFs. We think that the URLs add value to the references, particularly for data sets.}
  \end{itemize}

@@

    \resp{
      We have added two new imputation methods to the experiments (\Cref{sec:results}), to show that the planning algorithm can be used successfully with different imputation methods.
      We have also added a subsection to the algorithm section explaining how we chose the cost functions for each imputation method that we test and giving guidelines for choosing cost functions for a new imputation method (\Cref{sec:quality} and \Cref{sec:query-runtime}).
    }
  \end{itemize}

  \section{Reviewer 3}
  \begin{itemize}[leftmargin=*]
  \item W2. More work is required in the presentation: important parts are not explained in depth, while there are many details in other sections that can be omitted.

  \item W3. More experiments needed.

  \resp{
  We've extended the evaluation section to include more impute operators and an evaluation of planning times (\Cref{sec:results}).
  }

@@

  \item D1. Section 3, which is the meat of the paper, is very overloaded, without getting in depth in various parts of the system. It definitely should be split in more sections, and more details should be given. In particular:
    -- Give some more information about cardinality estimation. Also, Algorithm 3 is too far from the text, and needs some explanation too.
    -- Some more details are also required in the Imputation quality section. Why are you choosing these specific formulas for Loss, especially for the impute operator. Some examples would help.
    -- Algorithms 1 and 2 could probably be presented in an easier to grasp way.

    \resp{
    We've broken up this section and consolidated algorithm notation. The cardinality estimation section provides more detail about the algorithm. We also
    added intuition for how the heuristic \textsc{Penalty} (a renaming of Loss, for clarity) and \textsc{Time} should be designed. We walk through these definitions for three different
    impute operators (\Cref{sec:quality} and \Cref{sec:query-runtime}).
    }

@@

   Extending a rule-based optimizer to perform dynamic imputation would certainly be an interesting idea. However, the planning algorithm presented benefits from the tight
   interaction between traditional query planning and imputation operator placement. The latter can influence aspects such as join ordering and cardinality estimates for subplans.
   Incorporating these into a rule-based system would be possible, but we believe the plan selected through this approach will be suboptimal, compared to that chosen through
   our integrated planning approach.
   }

  \item D3. Is there any way for the user to know what error to expect based on the $\alpha$ value? Moreover, is its value somehow proportional to the error (e.g., $\alpha=0.5$ would give roughly double the error of $\alpha=0.25$)? Or should the user simply try multiple execution with different values of $\alpha$ until the desired results are given?

    \resp{In our intended usage, the user starts by running with $\alpha = 0$ and increases $\alpha$ if the query takes too long to complete.

     We agree that having a way of providing an estimate of error, based on $\alpha$, would be both useful and interesting to develop.
     Providing error estimates further requires a reasonable notion of ground-truth. For evaluation, we've used the values imputed on the 
     tables when applying whole-table imputation. Based on our evaluation, higher values of $\alpha$ correctly identify increased error
     within a query, however, extending that to be predictive is not something we implemented in the current system.
     }

@@

    \resp{We maintain the set of Pareto optimal plans at each point in the optimization. This set is significantly smaller than the set of all plans, so it represents some pruning (a point we have
    now clarified in the paper in \Cref{subsubsec:pareto}). We found that the size of these Pareto sets is small enough that we did not need to prune further.

      If necessary, it is possible to sample from the Pareto optimal plans during the plan enumeration. The final selected plan may not be Pareto optimal, but this is a potential way to control the number of plans.
      To perform this approximation, we only add plans to the frontier that are non-dominated and are sufficiently different from existing plans in the frontier. The plan already in the frontier should
      provide similar tradeoffs and reduces the search space explored. \ProjectName{} only employs this optimization when a large number of tables requiring imputation are involved in the query.
      }

    \item D7. When defining the cost of impute operator, how do you know a priori the number of iterations for the algorithm (given it has to converge)?

  \resp{For the iterative regression impute operator, we set a predefined number of iterations as an upper bound on its execution. If the imputation converges,
  we can return immediately, otherwise we return when this upper bound is reached. This is a common approach used in iterative optimization techniques, such
  as gradient descent. We've clarified this technique in the section discussing this imputation operation (\Cref{subsubsec:decision-trees}).
}

@@

    \resp{Yes. For example, in time series data, it is reasonable to fill the most recent non-null value forward. This kind of imputation isn't useful for survey data, so we didn't consider it for our experiments. We have extended the experiments section to consider a simple non-blocking imputation method relevant to our dataset: mean value imputation. This is a common technique, which fills in  missing values
    with the average observed value. In our case, we implement a non-blocking variant that estimates the mean from the plan's histograms, avoiding the need to perform a full pass over
    the data before imputing (\Cref{subsubsec:mean}).}

  \item D9. Additional experiments I suggest to include:
    -- Consider a scenario where the user submits multiple similar queries. Show after how many such queries your approach has no longer a competitive advantage compared to performing imputation on the data.
    -- Try different values for $\alpha$ (not just 0 and 1).
    -- Show runtime of enumeration algorithm as queries get more complex.

    \resp{
      It is straightforward to compute the number of queries before we lose our advantage.
      For example, performing full imputation on the ACS dataset takes 355 minutes (Sec. 5.3.2) and we can run a simple query in 4 seconds, so we can run approximately $5,000$ queries before taking more time than full imputation. We've made this point explicit in the ACS experiment section now.

      The sensitivity to different values of $\alpha$ depends on the query itself. Performing a scan of values of $\alpha$ with small step sizes can return
      a lot of the same plans for execution. We did run experiments with a range of $\alpha$ values, but instead of presenting a figure with each value of $\alpha$,
      which can result in many duplicate plans, we extended the results section to include the final frontiers for
      two queries from our benchmarks. This frontier can be exposed to the user to help in their workflow.

     The current set of results have also been extended to show an additional intermediate $\alpha$ value.

      We explored the planning times for queries with increasing number of joins by running additional experiments with up to 8 joins. 
      The standard algorithm presented in the original version submitted handles
      up to 5 joins, with columns requiring imputation in all tables, with planning times of approximately one second or less. We added a small optimization to help reduce
      planning times for a larger number of joins. We maintain an approximation of the Pareto sets, rather than the exact set, by only adding non-dominated plans that
      are sufficiently different. This optimization reduces planning time linearly with respect to the reduction in the size of the final frontier. Users can increase
      the level of approximation to reduce planning times further.}

### 2017-05-15 paper/rebuttal.tex

\setlist[itemize]{leftmargin=*,itemsep=-0.1em}

\begin{document}
\section{Introduction}
The authors would like to thank the reviewers for their helpful and insightful comments.
We have taken steps to address each comment and we believe that the paper is significantly improved
as a result. In this document, we give a point-by-point response to the comments from each reviewer.

In the rebuttal, the reviewer's original comments are in black and our responses are in \resp{blue}.
In our draft, new or significantly changed text is highlighted in {\color{blue}blue}.

\section{Overall changes}
\begin{itemize}[leftmargin=*]
\item \resp{
    Several reviewers commented on redundant figures, excessive whitespace, and
    formatting, so we've compressed many of the figures and removed whitespace to make room for additional material.
}

\item \resp{
    We have improved our system since the paper was submitted by adding imputation for categorical variables.
    This has a minor impact on the experimental results relative to the original submission.
    It improves the ability of the decision tree model to perform imputation by making these variables available to the model.
  }

@@

\item \resp{We extended experiments to include the new operators and provide more details on results for $\alpha$ (\Cref{sec:results}). We also added the Pareto frontiers for various plans (\Cref{fig:pareto-frontiers} ) and
explored planning times for more complex queries.}

\item \resp{We included a new section that provides a roadmap for implementing \ProjectName{} on top of an existing database (\Cref{sec:implementation}).}
\end{itemize}

  \section{Reviewer 1}
  \begin{itemize}[leftmargin=*]
  \item W2. The paper completely obscures the ML technique being used: (1) It would be nice to have some discussion of the complete API for implementing an ML technique, as well as how one might implement a cost-estimator for one or two examples. (2) It would be helpful to see the impact of a few different approaches (e.g., Tree-based, Interpolation, Regression, FD-based) on performance and accuracy. (3) Given that the ML time estimator is being treated as a black box, it would be useful to see how sensitive the plan selector is to inaccurate estimates.

  \resp{
    We extended our evaluation to include two commonly used imputation methods: mean value imputation and hot-deck imputation.
    These methods are in addition to the original imputation method which is based on decision trees.
    The decision tree model incorporates correlations between attributes, so should be able to take advantage of any functional dependencies to make better predictions.
    Interpolation-based imputation is useful for data where the ordering of tuples is significant, like time series, but is not well suited for survey data where the ordering of tuples is arbitrary.

@@

  \item S4.4.1: Imputation over the raw data is being used as ground truth. Is there any reason to suspect that this is the correct thing to do? Perhaps joining multiple tables together exposes new attributes that makes imputation *more* reliable? At the very least, it would be useful to understand why there is so much error in $\alpha=0$ on Q2 and Q8.

    \resp{For our purposes, imputation over the raw data is a reasonable ground truth. This
      is because the analyst, in practice, should be considering the tradeoffs on quality
      and performance of using ImputeDB versus a production imputation system on the raw
      data. Unfortunately, they would never have access to real ground truth in such a
      scenario, and moreover, finding ground truth for use even in an empirical evaluation
      of imputation methods can be challenging because artificially dirtying the data
      often does not capture the complex reasons for missingness in real world datasets.
      We updated the discussion in \Cref{sec:results} to emphasize this point.}

  \item There are a number of opportunities for trimming in the figures, tables, and algorithms: Algorithms 1-3 could be compacted a bunch, Table 1 is full of whitespace (e.g., make it a 3-column {figure*}), Tables 1-2 could also be summarized as CDFs, Figure 6 is full of whitespace, Table 3 could be put queries on a single row (e.g., Q1 would easily fit on 1 line), and the references have a lot of redundant information (URLs, for example).

    \resp{We have compacted the algorithms, tables, and references as possible. Note that \Cref{table:nhanes-description, table:fcc-description} show the percentage of missing values for \emph{each attribute}, not as an overall percentage, so cannot be summarized as CDFs. We think that the URLs add value to the references, particularly for data sets.}
  \end{itemize}

### 2017-05-15 paper/rebuttal.tex

  \item S4.4.1: Imputation over the raw data is being used as ground truth. Is there any reason to suspect that this is the correct thing to do? Perhaps joining multiple tables together exposes new attributes that makes imputation *more* reliable? At the very least, it would be useful to understand why there is so much error in $\alpha=0$ on Q2 and Q8.

    \resp{For our purposes, imputation over the raw data is a reasonable ground truth. This
      is because the analyst, in practice, should be considering the tradeoffs on quality
      and performance of using ImputeDB versus a production imputation system on the raw
      data. Unfortunately, they would never have access to real ground truth in such a
      scenario, and moreover, finding ground truth for use even in an empirical evaluation
      of imputation methods can be challenging because artificially dirtying the data
      often does not capture the complex reasons for missingness in real world datasets.
      We updated the discussion in \Cref{sec:results} to emphasize this point.}

  \item There are a number of opportunities for trimming in the figures, tables, and algorithms: Algorithms 1-3 could be compacted a bunch, Table 1 is full of whitespace (e.g., make it a 3-column {figure*}), Tables 1-2 could also be summarized as CDFs, Figure 6 is full of whitespace, Table 3 could be put queries on a single row (e.g., Q1 would easily fit on 1 line), and the references have a lot of redundant information (URLs, for example).

    \resp{We have compacted the algorithms, tables, and references as possible. Note that Tables 1 and 2 show the percentage of missing values for \emph{each attribute}, not as an overall percentage, so cannot be summarized as CDFs. We think that the URLs add value to the references, particularly for data sets.}
  \end{itemize}

### 2017-05-15 paper/introduction.tex
Update introduction of our baby

Many databases have large numbers of missing or \nullv{} values;  these can arise for a
variety of reasons, including missing source data,
missing columns during data integration, de-normalized databases, or as a result of outlier detection and cleaning where \nullv{} values are substituted for
bad values~\cite{kim2003}.  Such \nullv{} values can lead to incorrect or ill-defined query results~\cite{rubin1976}, and as such removing these values from data
before processing is often desirable.

% Solution
One common approach is to manually replace missing values using a statistical or predictive model based on other values in the table and record.
This process is called \emph{imputation}.
In this paper, we introduce \ProjectName{}, a new system designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.  \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}  
% Background
While traditional imputation methods work over the entire data set and replace all missing
values, running a sophisticated imputation algorithm over a large data set can be very
expensive: our experiments show that even a relatively simple decision tree algorithm takes \acsbaseresulthours{} to train and run
on a 600K row database.

>>>>

Many databases have large numbers of missing or \nullv{} values;  these can arise for a
variety of reasons, including missing source data,
missing columns during data integration, de-normalized databases, or outlier detection and
cleaning~\cite{kim2003}.  Such \nullv{} values can lead to incorrect or ill-defined query results~\cite{rubin1976}, and as such removing these values from data
before processing is often desirable.

% Solution
One common approach is to manually replace missing values using a statistical or predictive model based on other values in the table and record.
This process is called \emph{imputation}.
In this paper, we introduce new methods and theory to selectively applying
imputation to a subset of records, dynamically, during query execution.
We instantiate these ideas in a new system, \ProjectName{}.  \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}  
% Background
While traditional imputation methods work over the entire data set and replace all missing
values, running a sophisticated imputation algorithm over a large data set can be very
expensive: our experiments show that even a relatively simple decision tree algorithm takes \acsbaseresulthours{} to train and run
on a 600K row database.
